Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-16 index.qmd
section:issues; process:write

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory [-@Thurstone_1927b] posits that the dichotomous outcome resulting from comparing two stimuli is determined by two factors: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus. Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution. In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus. @fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).

@@

<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential. This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait. Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*. @fig-comparative_judgment shows an example distribution of the discriminal difference for two stimuli (objects).

@@

<!-- 3. Thurstone's Cases and their simplifying assumptions  -->
Importantly, the theory's general form primarily addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. Consequently, to enhance its practical applicability, Thurstone introduced five distinct cases, each defined by progressively simplifying assumptions. @fig-thurstone_cases summarizes these cases, focusing on key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersion across stimuli, the correlation between stimuli, and which judges perform the comparisons. For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#fig-thurstone_cases width=85%}

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Bramley_2008, pp. 254] (see @fig-thurstone_cases). This substitution has little effect on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (refer to @fig-logistic_vs_normal).

@@

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement. As a result, its assumptions may not be suitable for applications beyond the psychophysical contexts, which typically involve simple, homogeneous stimuli. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], further acknowledging that some assumptions could become problematic with less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretically or in practice: the assumption of zero correlation and equal dispersion between stimuli.

<!-- 3. The problems of assuming zero correlation between stimuli -->
[Regarding the assumption of *zero correlation between stimuli* ... ]{style="color:red;"}

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait under assessment [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

>>>>

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory [-@Thurstone_1927b] posits that the dichotomous outcome resulting from comparing two stimuli is determined by two factors: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus. Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution. In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the stimulus' perceived trait level. @fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).

@@

<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
However, since the discriminal mode and dispersion of a single stimulus are not directly observable except through comparison, the *law of comparative judgment* becomes essential. This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait. Thus, the observed dichotomous outcome is determined by the distribution of the difference between the stimuli's discriminal processes, called the *discriminal difference*. @fig-comparative_judgment shows an example distribution of the discriminal difference for two stimuli (objects).

@@

<!-- 3. Thurstone's Cases and their simplifying assumptions  -->
Importantly, the theory's general form primarily addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. Consequently, to enhance its practical applicability, Thurstone introduced five distinct cases, each defined by progressively simplifying assumptions. @tbl-thurstone_cases summarizes these cases, focusing on key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersions across stimuli, the correlation between stimuli, and which judges perform the comparisons. For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little effect on the model's estimation or interpretation, as the Normal and Logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (refer to @fig-logistic_vs_normal).

@@

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement. As a result, its assumptions may not suit applications beyond the psychophysical contexts for which it was created. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic with more complex, less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.

<!-- 3. The problems of assuming zero correlation between stimuli -->
[The *zero correlation between stimuli* represents two additional assumtions about the method:   ]{style="color:red;"}

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait of interest [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying texts' quality level (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].
